\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-02-28 paper/main.tex
delete old comments
finish conclusion

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on perfect memory payoffs and limited memory payoffs. In the case of
perfect memory payoffs we use Eq~\ref{eq:perfect_memory_lambdas} when we
estimate the probability that a mutant fixes, and
Eq~\ref{eq:limited_memory_lambdas} in the case on limited memory. We run two
independent runs for each approach and we vary the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} shows the simulation
results. It depicts the evolving conditional cooperation probabilities $p$ and
$q$. The discount factor~$\delta$ is comparably high, thus we do not report the
opening move \(y\) as it is a transient effect. The upper panel corresponds to
the standard scenario considered in the literature, it considers players who use
expected payoffs to update their strategies. The bottom panel shows the scenario
considered herein, in which players update their strategies based on their last
round's payoff.

The figure suggests that when updating is based on perfect memory payoffs
players tend to be more generous and more cooperative. The $q$-values of the
resident strategies are on average higher in the case of perfect memory. Thus,
players will occasionally forgive a defection more often if their fitness
depends on interacting with every member of the population, and this effect
becomes more notable as the value of the benefit increases. In the case of
limited memory the resident strategies are less forgiving and their forgiveness
is independent of the values of benefit. In the Supplementary Information
Sections 2.1 and 2.2.1 we show that $q$-values of a resident strategy have to be less
than \(1 - c/b\) in the perfect memory case and less than \(0.5\) in the limited
memory case. The generosity for the perfect memory case is always higher for a
benefit value higher than 2, and it increases as the benefit value increases.
This leads to a more cooperative population. We calculate the average
cooperation rate for each simulation which is the average cooperation rate
within the resident population. In the case of the perfect memory payoffs the
average cooperation rate is strictly higher than that of the last round payoffs,
and the difference is most for $b=10$. The average cooperation of resident
strategies drops from 97\% to 57\%. This indicates that the expected payoffs
overestimate the evolved cooperation.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    ({\bf Schematic illustrations}) On the left panels we show schematic
    illustrations of the perfect memory and the limited memory cases. The shaded
    background denotes the game phase information that an individual considered
    when updating strategies. In the case of perfect memory the entire region is
    shaded and in the case of limited memory only one turn with a single member
    of the population. ({\bf Simulations}) We have run four simulations of the
    pairwise comparison process for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since
    simulations are run for a relatively high continuation probability of
    $\delta\!=\!0.999$, we do not report the players' initial cooperation
    probability $y$. The graphs show how often the resident population chooses
    each combination ($p,q$) of conditional cooperation probabilities in the
    subsequent rounds. We also report the evolved cooperation rate which is
    calculated as the average cooperation rate within the resident population.
    ({\bf Perfect Memory}) In the case of low benefit the resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. Conditional cooperators, or otherwise known as
    generous tit for tat, are a set of strategies that always cooperate
    following a cooperation ($p\!\approx\!1\!$) and cooperate with a probability
    $q$ given that the co-player has defected. $q$ denotes the generosity of a
    player. The resident population applies a conditional cooperator strategy
    for which $q\!\le\!1\!-\!c/b\!=\!0.67$. This is true also for the case of
    high benefit. In this case the population mainly consists of conditional
    cooperators of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!1/10\!=\!0.9$). In
    the Supplementary Information Section 2.1 we show that a conditional
    cooperator needs to be of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!c/b$) to
    not be invaded by defecting strategies. A higher generosity in the
    population results in a higher average cooperation rate. The average
    cooperation rate increases from 52\% for \(b\) to 98\% for \(b=10\). ({\bf
    Limited Memory}) When players update their strategies based on their
    realized payoffs in the last round, there are two different predominant
    behaviors regardless of the benefit value. The resident population either
    consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of conditional
    cooperators. The maximum level of $q$ consistent with stable cooperation is
    somewhat smaller compared to the perfect memory setting, $q\!<\!0.5$.
    Namely, in the Supplementary Information Section 2.2.1 we show that
    regardless of the value of benefit, in the case of limited payoff memory, an
    individual needs a generosity smaller of $0.5$ to repel defectors. The plots
    are fairly similar regardless of the benefit, and the evolved cooperation
    rate only slightly increases from 37\% to 51\%. Parameters: $N\!=\!100$,
    $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effects of the cooperation benefit and the strength of
selection on the evolved generosity \(q\) and cooperation rate
(Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}).
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A} suggests that
perfect memory always yields a higher cooperation rate. We observe that the
cooperation rate increases as the value of the benefit gets higher, whereas in
comparison for the limited memory payoffs, the cooperation rate remains
unchanged (as we would have expected) at approximately 50\% once \(b=5\). From
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B} we observe
that for weak selection, \(\beta < 1\), the two methods yield similar results,
however, as \(\beta\) increases there is variation in the evolving populations.
In the case of expected payoffs the resident populations become more cooperative
as \(\beta\) increases, whereas in the case of limited memory payoffs, the
resident populations become more defective.

@@

\subsection{Updating Payoffs based on More Memory}

So far we have explored the difference between the expected payoffs and the last
round payoffs. To explore further the effect of limited memory we consider three
more cases. These are the cases where the updating payoffs depend on  the last
round with two members of the population, the last two rounds with another
member of the population, and the last two rounds with two members of the
population. Similar to the last round payoff. The details of each approach can
be found in the Supplementary Information. We use numerical simulation of the
pairwise process for the various schemes. The results for low and high benefit
are shown in Figure~\ref{fig:cooperation_rate_all_updating_payoffs}.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different updating payoffs.}
  From left to right, we present result on the following updating payoffs cases;
  (1) the expected payoffs (perfect memory), (2) the last round payoff from one
  interaction (limited memory), (3) the last round payoff from two interactions,
  (4) the last two rounds payoffs from one interaction, (5) the last two rounds
  payoffs from two interactions. The details of each case are given in the
  Supplementary Information. Cases 3 and 4 have similar results. In that the
  evolved generosity is the same. For cases 2 - 5 the cost benefit ratio
  has an effect compared to the limited memory case. The cooperation rate
  increases, and as we allow for more information the closer we move to the
  perfect memory. Unless explicitly varied, the parameters of the simulation are
  $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

\newpage

@@

A major focus of evolutionary game theory is the evolution of cooperation among
selfish individuals. Sometimes evolutionary game theory even seems to be reduced
to the evolution of cooperation [Arne2022]. “Why do we act altruistically to
provide a benefit to others when there is a cost to ourselves?” is still a
puzzling question amongst researchers [Glynatsi]. In this determinist
formulation, a small fraction of cooperative individuals that have a lower
fitness can never invade a population of defecting strategies.  The stochastic
effect can lead to fundamental changes since disadvantageous mutants have a
small yet non-zero probability to reach fixation in a finite population, and
thus cooperation can emerge.

Cooperation can be seen at odd, why is it that we choose to help others,
increasing their payoff, at the expense of decreasing one's own? In spite of all
the selfish genes' animal and human communities seem to altruistically help each
other and cooperate, and evolutionary game theory has helped us shape our
understanding of the evolution of cooperation.

Previous evolutionary models often feature a curious inconsistency. When
modeling how individuals make decisions in each round, these models assume that
players only remember the last round. However, when modeling how individuals
update their strategies over time, individuals are assumed to have perfect
memory.

Here, we have explored how robust cooperation is as models deviate from the
perfect memory assumption. Initially we considered the donation game. We showed
that when the last round payoff is used instead of the expected payoffs,
cooperation can even evolve when individuals only use a minimum of information,
however, the evolving cooperation rates are typically lower. The resident
strategies were both less cooperative and less generous. This effect was only
intensified we increase the benefit and the strength of selection independently.
The results showed that as each parameter was being increased the difference
between the cooperation rates were widening. This indicates that cooperative
players benefit from being able to interact with everyone in the population.

We extended our approach and presented results not only based on different
classes of social games, but also based on different limited memory payoffs. The
analysis showed that specifically for the case of the prisoner's dilemma games
the contrast between the two approaches can be significantly large. The
prisoner's dilemma is one of the most applied types of social games when we
discuss the evolution of cooperation. Our results indicate that cooperation
struggles to evolve in the prisoner's dilemma when only a minimum of social
information is used at the updating stage.

@@

% \appendix

% \section{Model Setup}\label{appendix:methods}

% Consider a population of \(N\) individuals where \(N\) is even. At any point in
% time there are at most two different strategies in present in the population.
% More specifically, a mutant strategy played by \(k\) individuals and a resident
% strategy played by \(N - k\) individuals. We assume a pairwise process in which
% strategies spread because they are imitated more often. Each step of the
% evolutionary process consists of two stages; a game stage and an update stage.

% In the game stage, each individual is randomly matched with some other
% individual in the population. Their interaction lasts for a number of turns
% which is not fixed but depends on the continuation probability \(\delta\). At
% each turn the individuals choose between cooperation (\(C\)) and defection
% (\(D\)). Thus, there are four possible outcomes in each turn \(CC, CD, DC\) and
% \(DD\). If both players cooperate they receive the reward payoff \(R\), whereas
% if both players defect they receive the punishment payoff \(P\). If one
% cooperates but the other defects, the defector receives the temptation to
% defect, \(T\), whereas the cooperator receives the sucker's payoff, \(S\). Let
% $\mathcal{U}=\{R,S,T,P\}$ denote the set of feasible payoffs in each round, and
% let $\mathbf{u}\!=\!(R,S,T,P)$ be the corresponding payoff vector. We present
% results for various values of $\mathcal{U}$ for all the symmetric \(2 \times 2\)
% games.

% A further assumption of our model is that individuals make use of reactive
% strategies when they make decisions in each round. Reactive strategies are a set
% of strategies that take into account only the previous action of the opponent. A
% reactive strategy can be written explicitly as a vector,

% \[s=(y, p, q)\]

% where \(y\) is the probability that the strategy opens with a cooperation and
% \(p, q\) are the probabilities that the strategy cooperates given that the
% opponent cooperated and defected equivalently.

% In the updating stage, two players are randomly drawn from the population, a
% `learner' and a `role model'. The learner adopts the role model's strategy based
% on the Fermi distribution function, %ToDo add reference

% % \begin{equation} \label{Eq:rho}
% % \rho(u_{L}, u_{RM}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (u_{RM}\!-\!u_{L})}}.
% % \end{equation}

% where $u_{L}\!\in\! \mathcal{U}$ is the learner's payoff, $u_{RM}\!\in\!
% \mathcal{U}$ is the role model's payoff, and $\beta\!\ge\!0$ is the strength of
% selection.

% We iterate this basic evolutionary step until either the mutant strategy goes
% extinct, or until it fixes in the population and becomes the new resident
% strategy. After either outcome, we set $k$ to $1$ and we introduce a new mutant
% strategy which is uniformly chosen from all reactive strategies at random.
% Instead of simulating each step of the evolutionary process, we estimate the
% probability that a newly introduced mutant fixes~\cite{nowak2004emergence}. This
% is defined as the fixation probability of the mutant, and the standard form is
% the following,

% \begin{equation}\label{eq:appendix_fixation_probability}
% \varphi = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}},
% \end{equation}

% where \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
% mutants decreases and increases respectively.

% This process of mutation and fixation/extinction is iterated many times. The
% evolutionary process is summarized by
% Algorithm~\ref{algorithm:pairwise_comparison}.

% \begin{algorithm}[!htbp]
%   \SetAlgoLined $N \leftarrow$ population size\; $k \leftarrow 1$\; resident
%    $\leftarrow (0, 0, 0)$\; \While{ $t <$ maximum number of steps}{mutant
%    $\leftarrow$ random: $\{\emptyset \}\ \rightarrow R^3$\; fixation probability
%    $\leftarrow \varphi $\; \If{$\varphi >$ random: $i \rightarrow [0,
%    1]$}{resident $\leftarrow$ mutant;}} \caption{Evolutionary
%    process}\label{algorithm:pairwise_comparison}
% \end{algorithm}

% The aim of this work is to explore the effect of updating memory on the
% cooperation rate of the evolved population. For this reason we consider two
% different approaches when estimating the payoffs at the updating stage. The two
% approaches we consider are those of (i) the expected and (ii) the limited memory
% payoffs.

% \subsection*{Expected Payoffs}

% The expected payoffs are the conventional payoffs used in the updating
% stage~\cite{imhof2010stochastic}. They are defined as the mean payoff of an
% individual in a well-mixed population that engages in repeated games with all
% other population members.

% We first define the payoff of two reactive strategies at the game stage. Assume
% two reactive strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$.
% It is not necessary to simulate the play move by move, instead the play between
% the two strategies is defined a Markov matrix \(M\),

% \begin{equation}\label{eq:transition_matrix}
%   \input{static/matrix.tex}.
% \end{equation}

% whose stationary vector \(\mathbf{v}\), combined with the payoff \(u\), yields
% the game stage outcome for each strategy,
% \(\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle\)~\cite{Hauert1997}.

% In the updating stage the learner adopts the strategy of the role model based on
% their updating payoffs. Given that there are only two different types in the
% population at each time step we only need to define the expected payoff for a
% resident (\(\pi_R\)) and for a mutant (\(\pi_M\)). Assume the resident strategy
% \(s_R = (y_R, p_R, q_R)\) and the mutant strategy \(s_M = (y_M, p_M, q_M)\), the
% expected payoffs are give by,

% \begin{equation} \label{Eq:ExpPay}
%   \begin{array}{lcrcr}
%   \displaystyle \pi_R	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(s_R,s_R),\mathbf{u}\rangle	&+	&\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(s_R,s_M),\mathbf{u}\rangle,\\[0.5cm]
%   \displaystyle \pi_M	&=	&\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(s_M,s_R),\mathbf{u}\rangle&+	&\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(s_M,s_M),\mathbf{u}\rangle.\\
%   \end{array}
% \end{equation}

% The number of mutant in the population increase if a learner resident adopts the
% strategy of a mutant role model, and decreases if a mutant leaner adopts the
% strategy of a resident. The probabilities that the number of mutants decreases
% and increases, \(\lambda^-_k\) and \(\lambda^+_k\), are now explicitly defined
% as,

% \begin{align*} 
%   \lambda^-_k &\!=\!\rho(\pi_R, \pi_M) \\
%   \lambda^+_k &\!=\!\rho(\pi_M, \pi_R).
% \end{align*}

% \subsection*{Limited memory payoffs}

% Initially, we discuss the case of the \textbf{last round updating payoff}. At
% the stage game we define the payoff of a reactive strategy in the last round,
% Proposition~\ref{proposition:last_round}.

% \begin{Prop}\label{proposition:last_round} Consider a repeated game, with
%     continuation probability $\delta$, between players with reactive strategies
%     $s_1\!=\!(y_1, p_1, q_1$)  and $s_2\!=\!(y_2,p_2,q_2)$ respectively. Then
%     the probability that the $s_1$ player receives the payoff $u\!\in\!
%     \mathcal{U}$ in the very last round of the game is given by
%     $v_{u}(s_1,s_2)$, as given by Equation~(\ref{Eq:LastRound}).

%     \begin{equation} \label{Eq:LastRound}
%       \setlength{\arraycolsep}{1pt}
%       \begin{array}{rcl}

%       v_{R}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
%       {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times R,\\[1cm]

%       v_{S}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
%       {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times S,\\[1cm]

%       v_{T}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
%       {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times T,\\[1cm]

%       v_{P}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
%       {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times P.
%       \end{array}
%     \end{equation}

% In these expressions, we have used the notation $r_i:=p_i\!-\!q_i$,
% $\bar{y}_i\!=\!1\!-\!y_i$, $\bar{q}_i:=1\!-\!q_i$, and
% $\bar{r}_i:=\bar{p}_i\!-\!\bar{q}_i=-r_i$ for $i\!\in\!\{1,2\}$.
% \end{Prop}

% \begin{proof}
% Given a play between two reactive strategies with continuation probability
% $\delta$. The outcome at turn \(t\) is given by,

% \begin{equation}\label{eq:}
%   (1 - \delta) \mathbf{v_0} \sum \delta^{t} M^{(t)},
% \end{equation}

% where $\mathbf{v_0}$ denotes the expected distribution of the four outcomes in
% the very first round, and \(1- \delta\) the probability that the game ends. It
% can be shown that,

% \begin{align*}
%   (1 - \delta) \mathbf{v_0} \sum \delta^{t} M^{(t)} & = (1 - \delta)(\mathbf{v_0} + \delta \mathbf{v_0} M + \delta^{2}\mathbf{v_0} M ^{2} + \dots )\\ 
%    & = (1 - \delta)\mathbf{v_0} (1 + \delta M + \delta^{2}M ^{2} + \dots ) \text{ using standard formula for geometric series}\\ 
%    & = (1 - \delta)\mathbf{v_0}(I_4 - \delta M)^{-1}
% \end{align*}

% where \((1 - \delta)\mathbf{v_0}(I_4 - \delta M)^{-1}\) is vector \(\in R^{4}\)
% and it the probabilities for being in any of the outcomes \(CC, CD, DC, DD\) in
% the last round. Combining this with the payoff vector \(u\) and some algebraic
% manipulation we derive to the Equation~\ref{Eq:LastRound}.
% \end{proof}

% In the updating stage we select a mutant and resident to be either the role
% model or the learner. Given that they can interact with only one other member of
% the population, they can interact either with each other or either can interact
% with another resident or with another mutant. Thus, in each updating stage there
% are five possible combinations of pairs (Figure~\ref{fig:single_pairs}).

% Given the last round payoff and possible pair combinations for a single
% interaction, we define the probability that the respective last round payoffs of
% two players \(s_1, s_2\) are given by $u_1$ and $u_2$ as,

% \begin{equation}\label{eq:Chi}
% \setlength{\arraycolsep}{1pt}
% \begin{array}{llrl}
% x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
% &+	
% &\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
% &\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(s_1,s_2) v_{u_2}(s_2,s_2) + 
%  \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(s_1,s_2) v_{u_2}(s_2,s_1)\right.\\[0.5cm]
% &&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(s_1,s_1) v_{u_2}(s_2,s_2) + 
%  \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(s_1,s_1) v_{u_2}(s_2,s_1)\right].
% \end{array}
% \end{equation}

% The first term on the right side corresponds to the case that the learner and
% the role model happened to be matched during the game stage, which happens with
% probability $\frac{1}{(N\!-\!1)}$. In that case, we note that only those payoff
% pairs can occur that are feasible in a direct interaction,
% $\splitatcommas{(u_1,u_2)\in \mathcal{U}^2_F:=\big\{ (R,R), (S,T), (T,S), (P,P)
% \big\}}$, as represented by the respective indicator function. Otherwise, if the
% learner and the role model did not interact directly, we need to distinguish
% four different cases, depending on whether the learner was matched with a
% resident or a mutant, and depending on whether the role model was matched with a
% resident or a mutant.

% Given that $N\!-\!k$ players use the resident strategy
% $s_{R}\!=\!(y_{R},p_{R},q_{R})$ and that the remaining $k$ players use the
% mutant strategy $s_{M}\!=\!(y_{M},p_{M},q_{M})$, the probability that the number
% of mutants increases by one in one step of the evolutionary process can be
% written as

% \begin{align}
% \lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M})\cdot \rho(u_{R},u_{M}), \\
% \lambda^-_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M})\cdot \rho(u_{M},u_{R}).
% \end{align}

% In this expression, $\frac{(N\!-\!k)}{N}$ is the probability that the randomly
% chosen learner is a resident, and $\frac{k}{N}$ is the probability that the role
% model is a mutant. The sum corresponds to the total probability that the learner
% adopts the role model's strategy over all possible payoffs $u_R$ and $u_M$ that
% the two player may have received in their respective last rounds. We use
% $x(u_R,u_M)$ to denote the probability that the randomly chosen resident
% obtained a payoff of $u_R$ in the last round of his respective game, and that
% the mutant obtained a payoff of $u_M$.

% We extend our framework to consider the case where players update their
% strategies based on the outcome of \textbf{the last two turns and based on their
% interaction with two other members of the population}. At the stage game we
% define the payoff of a reactive strategy in the last two rounds,
% Proposition~\ref{proposition:last_two_rounds}.

% \begin{Prop}\label{proposition:last_two_rounds} Consider a repeated game, with
%   continuation probability $\delta$, between players with reactive strategies
%   $s_1\!=\!(y_1, p_1, q_1$)  and $s_2\!=\!(y_2,p_2,q_2)$ respectively. Let
%   $\mathcal{\tilde{U}} = \splitatcommas{\{RR, RS, RT, RP, SR, SS, ST, SP, TR,
%   TS, TT, TP, PR, PS, PT, PP\}}$ denote the set of feasible payoffs in the last
%   two rounds, and let \(\tilde{\mathbf{u}}\) be the corresponding payoff vector.
%   Then the probability that the $s_1$ player receives the payoff $u\!\in\!
%   \mathcal{\tilde{U}}$ in the very last two rounds of the game is given by,

%   \begin{equation}
%   \langle\mathbf{\tilde{v}}(s_1,s_2),\mathbf{\tilde{u}}\rangle, \text{ where } \mathbf{\tilde{v}} \in R^{16} \text{ is given by },
%   \end{equation}
%   \begin{equation}
%     \mathbf{\tilde{v}}(s_1,s_2) = (1 - \delta) m_{a_1, a_2} \delta^2 \left[\mathbf{v_0}(I_4 - \delta M)^{-1}\right]_{a_1, a_2}, \quad  m_{a_1, a_2} \in M \ \forall \ a_1, a_2 \in \{1, 2, 3, 4\}.
%   \end{equation}
% \end{Prop}

% In the updating stage we select a mutant and resident to be either the role
% model or the learner. Given that they can interact with two other members of the
% population there are a total of twenty four possible combinations of pairs
% (Figure~\ref{fig:pissible_two_pairs}).

>>>>

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on perfect and limited payoff memory. We performed two independent runs
for each approach varying the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (since the discount
factor~$\delta$ is comparably high we do not report the opening move \(y\)). The
upper panel of Figure~\ref{fig:expected_and_stochastic_for_donation} corresponds
to the standard scenario considered in the literature, and the bottom panel
shows the scenario considered herein, in which players update their strategies
based on their last round's payoff.

The figure suggests that when updating is based on perfect payoff memory players
tend to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in this case, thus, players will occasionally
forgive a defection more often if their fitness depends on interacting with
every member of the population. This effect becomes more notable as the value of
the benefit increases. In the case of limited memory the resident strategies are
less forgiving and their forgiveness regardless of the values of benefit. In
analysis carried out in the Supplementary Information Section 3 we show that a
resident population of conditional cooperators with \(q > \frac{1}{2}\) can not
avoid invasion by a defecting strategy. This holds independently of the cost and
benefit values.

Higher $q$ values lead to a more cooperative population. We calculate the
average cooperation rate for each simulation which is the average cooperation
rate within the resident population. In the case of the perfect payoff memory 
the average cooperation rate is strictly higher than that of the last round
payoffs. The difference between the cooperation rate is the highest at $b=10$.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.9\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    ({\bf Schematic illustrations}) On the left panels we show schematic
    illustrations of the perfect memory and the limited memory cases. The shaded
    background denotes the game phase information that an individual considers
    when updating strategies. In the case of perfect payoff memory the entire region is
    shaded and in the case of limited payoff memory only one turn with a single member
    of the population. ({\bf Simulations}) We have run four simulations of the
    pairwise comparison process for $T\!=\!10^7$ time steps. For each time step
    of the process
    we record the current resident population ($y,p,q$). Since
    simulations are run for a relatively high continuation probability of
    $\delta\!=\!0.999$, we do not report the players' initial cooperation
    probability $y$. The plots show how often the resident population chooses
    each combination ($p,q$) of conditional cooperation probabilities in the
    subsequent rounds. We also report the evolved cooperation rate which is
    calculated as the average cooperation rate within the resident population.
    ({\bf Perfect Memory}) In the case of low benefit the resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. Conditional cooperators, or otherwise known as
    generous tit for tat, are a set of strategies that always cooperate
    following a cooperation ($p\!\approx\!1\!$) and cooperate with a probability
    $q$ given that the co-player has defected. $q$ denotes the generosity of a
    player. The resident population applies a conditional cooperator strategy
    for which $q\!\le\!1\!-\!c/b\!=\!0.67$. 
    In the case of high benefit the population mainly consists of conditional
    cooperators of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!1/10\!=\!0.9$). In
    the Supplementary Information Section 2 we show that a conditional
    cooperator needs to be of the form ($p\!\approx\!1\!, q\!\le\!1\!-\!c/b$) to
    not be invaded by defecting strategies. A higher generosity in the
    population results in a higher average cooperation rate. The average
    cooperation rate increases from 52\% for \(b=3\) to 98\% for \(b=10\). ({\bf
    Limited Memory}) When players update their strategies based on their
    realized payoffs in the last round, there are two different predominant
    behaviors regardless of the benefit value. The resident population either
    consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of conditional
    cooperators. The maximum level of $q$, consistent with stable cooperation, is
    somewhat smaller compared to the perfect memory setting.
    Namely, in the Supplementary Information Section 3 we show that
    regardless of the value of benefit a conditional cooperator need to be of
    the form $q\!<\!\frac{1}{2}$ to not be invaded by defectors.
    The evolved cooperation
    rate only slightly increases from 37\% ($b=3$) to 51\% ($b=10$). Parameters: $N\!=\!100$,
    $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effects of the cooperation benefit and the strength of
selection on the evolved generosity \(q\) and on the cooperation rate
(Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}).
Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{A} suggests that
perfect memory always yields a higher cooperation rate. In this case the
cooperation rate increases as the benefit increases. In comparison for limited
payoff memory the cooperation rate remains unchanged at approximately 50\% once
\(b=5\).  From Figure~\ref{fig:cooperation_rate_over_benefit_and_beta}\textbf{B}
we observe that for weak selection, \(\beta < 1\), the two methods yield similar
results, however, as \(\beta\) increases there is variation in the evolving
populations. In the case of expected payoffs the resident populations become
more cooperative, whereas in the case of limited payoff memory, the resident
populations become more defective.

@@

Up to this point we have discussed only two edge cases. The framework of limited
payoff memory can be easily extended to allow individuals to observe more
rounds, to interact with more members or both (see Supplementary Information
Sections 4-6). To further understand the effect of limited payoff memory we
consider the cases of updating based on the last round with two members of the
population, the last two rounds with another member of the population, and the
last two rounds with two members of the population. Similar to the last round
payoff. We perform numerical simulations for each of the fitness approaches. In
Figure~\ref{fig:cooperation_rate_all_updating_payoffs} we report the cooperation
rates for low and high values of benefit.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different updating payoffs.}
  From left to right, we present result on the following updating payoffs cases;
  (a) the expected payoffs (perfect memory), (b) the last round payoff from one
  interaction (limited memory), (c) the last round payoff from two interactions,
  (d) the last two rounds payoffs from one interaction, (e) the last two rounds
  payoffs from two interactions. For the updating payoffs (c) and (d) we have
  carried out an analysis which has shown that conditional cooperators should
  adopt a \(q\) value smaller than $\frac{\delta - 1 +
  \frac{\sqrt{2}}{2}}{\delta}$ and $\frac{\delta + \sqrt{\delta^{2} + 1} - 1}{2 \delta}$
  respectively. Note that as \(\delta
  \rightarrow 1\) both right hand sides tend to \(\frac{\sqrt{2}}{2}\).
  Regardless the cooperation rate between for case (c) is slightly higher.
  In case (d) the cooperation rate is the second highest hinting that as we
  allow for more information the closer we move to the perfect payoff memory.
  Unless explicitly varied, the parameters of the simulation are
  $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
  Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

@@

Cooperation can be seen at odd, why is it that we choose to help others at a
personal cost? In spite of all the selfish genes', animal and human communities
show signs of altruism and cooperation~\cite{milinski1987tit, kerr2002local,
carter2020development}. Evolutionary game theoretical models have helped us
shape our understanding of the evolution of cooperation. In fact, the evolution
of cooperation constitutes such a major focus of the field that evolutionary
game theory seems to be reduced to the evolution of
cooperation~\cite{Traulsen:PhilTrans:2022}. Previous evolutionary models often
feature a curious inconsistency. When modeling how individuals make decisions in
each round, these models assume that players only remember the last round.
However, when modeling how individuals update their strategies over time,
individuals are assumed to have perfect memory.

Here, we have explored how robust cooperation is as models deviate from the
perfect memory assumption. We considered a special case of the prisoner's
dilemma, the donation game. The prisoner's dilemma is one of the most applied
types of social games when we discuss the evolution of cooperation. We showed
that when the last round payoff is used instead of the expected payoffs,
cooperation struggles to evolve  when only a minimum of social information is
used at the updating stage. The resident strategies were both less cooperative
and less generous. This effect was only intensified we increase the benefit and
the strength of selection independently. The results showed that as each
parameter was being increased the difference between the cooperation rates were
widening. This indicates that cooperative players benefit from being able to
interact with everyone in the population.

We extended our approach and presented results on different limited memory
payoffs. The analysis showed cooperation can even evolve when we allow
individual to consider more information.

@@
